Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-05-28 paper/main.tex
intro

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}
\note{please prioritize flagging content that needs major revisions: (1) Wrong (2) Missing (3) Horrific  (4) Incoherant (5) Movable to appendix - TCVG page limit is 12 pages content + supplemental}

\IEEEPARstart{V}isualizations are expected to be a faithful translation of the inherent structure of the data being visualized; therefore the components of visualization software libraries that translate data to graphics are also expected to preserve this structure. Over time, library developers may need to develop new components to adapt to changing data needs, especially high dimensional multivariate and distributed data. 
This paper provides a method for formally describing structure that should be preserved and this work provides guidance on how to verify this preservation. This abstraction of the data structure and specification of the components can then be used to guide the developement of functional\footnote{functional programming paradigm} components. Since these components are functional, they are expected to be more composable and modular than existing components, have fewer side effects and therefore be easier to incorporate alignside existing library elements, and be verifiable and therefore easier to test \cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

\section{Related Work}
We present an abstraction of data that is agnostic to the types of fields or number of dimensions of the data by building on work that has codified structure and how it is preserved. Structure is mainly described as a combination of the topological properties of the data and the mathematical structure of the dataset fields. Structure preservation is expressed as equivalent changes on the data input and graphics output of a visualization algorithm. Our model encapsulates all these notions of structure by providing a language for expressing topology, field structure, and changes to the data model.

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are; expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fbtau.png}
  \caption{A non-trivial fiber bundle is bundle where the (by definition isomorphic) fibers change along the indexing base space. Here $\dfiber_{\dbasepoint_1}$ and $\dfiber_{\dbasepoint_2}$ go in reverse directions and the global section $sin$ returns $sin(\dbasepoint_{i})$ which is a value between $[0,1]$
  \label{fig:atct:stripbundle}}
\end{figure}

@@

An artist is equivariant when changes to expected and actual are equivariant. As introduced in \autoref{eq:atct:morphism:base}, the base space transformation \dfunch\ is invariant because $\dsection\restriction_{\openset} = \dsection\restriction_{\dfunch(\openset^{\prime})}$. This means that, for all points in the data $\dbasepoint \in \dbase$, the measurement should not change if only the base space is transformed 
\begin{equation}
  \label{eq:atct:equivarance:verify:base}
  \equivc(\dsectionc)(\dfunchc(\dbasepointc^{\prime})) = \extractmc(\vartistc(\dsectionc))(\dbasepointc)
\end{equation}
On the other hand, a change in sections \autoref{eq:atct:morphism:fiber} induces an equivalent change in measurements
\begin{equation}
  \label{eq:atct:equivarance:verify:fiber}
  \equivc(\dfunctc(\dsectionc))(\dbasepointc) = \dfunctc_{\measure}(\extractmc(\vartistc(\dsectionc))(\dbasepoint))
\end{equation}
The change in measurements $\dfunctc_{\measure}$ is defined by the developer as the symmetry between data and graphic that the artist is expected to preserve.

@@

\subsubsection{Addition: +}
\label{sec:artist:addition}
As illustrated in \autoref{eq:artist:operator}, data bundles can be combined by taking the disjoint  union of base spaces $\dbase^{a} \sqcup_{\dbase^c} \dbase^b$, where $\dbase^c$ is an overlap. When the fibers of each bundle are isomorphic $\dfiber^{a} \simeq \dfiber^{b}$, which we denote as $\dtotal$, this is analogous to adding more observations to a dataset. We propose an addition operator that states that an artist that takes in a dataset can be constructed using artists that take as inputs subsets of the dataset

>>>>

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}
\note{need a math check + rough cut on sections 4+}

\IEEEPARstart{V}isualizations are expected to be a faithful translation of the inherent structure of the data being visualized; therefore the components of visualization software libraries that translate data to graphics are also expected to preserve this structure. Motivated by a need to develop new library components that adapt to modern data needs, especially high dimensional multivariate and distributed data, we propose a framework that provides a language to express that topological and field structure of the input data that must be mapped to the visualization and which changes to data must be expressed in visualization, and a functional framework that describes composability in terms of functional algebraic operations. This allows complex visualizations to be built from simple verifiable parts\cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}, which makes these components easier to drop into the existing library and be used by domain specific library developers to build new libraries that work with existing ones.  While many of these elements have been partially described elsewhere, our framework naturally combines them in a unified framework based on ideas in category theory. Further we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

\section{Related Work}
In this work, we propose a uniform method for expressing the structure that visualizations are expected to preserved by combining and generalizing previous work that describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable, reusable software library components with more consistent interfaces.

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are; where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

\begin{figure}[h!]
  \centering
  %needs cropping
  \includegraphics[scale=.5]{fbtau.png}
  \caption{A non-trivial fiber bundle is bundle where the (by definition isomorphic) fibers change along the indexing base space. Here $\dfiber_{\dbasepoint_1}$ and $\dfiber_{\dbasepoint_2}$ go in reverse directions and the global section $sin$ returns $sin(\dbasepoint_{i})$ which is a value between $[0,1]$
  \label{fig:atct:stripbundle}}
\end{figure}

@@

An artist is equivariant when changes to the input and output are equivariant.

As introduced in \autoref{eq:atct:morphism:base}, the base space transformation \dfunch\ is invariant because $\dsection\restriction_{\openset} = \dsection\restriction_{\dfunch(\openset^{\prime})}$. This means that, for all points in the data $\dbasepoint \in \dbase$, the measurement should not change if only the base space is transformed 
\begin{equation}
  \label{eq:atct:equivarance:verify:base}
  \equivc(\dsectionc)(\dfunchc(\dbasepointc^{\prime})) = \extractmc(\vartistc(\dsectionc))(\dbasepointc)
\end{equation}
On the other hand, a change in sections \autoref{eq:atct:morphism:fiber} induces an equivalent change in measurements
\begin{equation}
  \label{eq:atct:equivarance:verify:fiber}
  \equivc(\dfunctc(\dsectionc))(\dbasepointc) = \dfunctc_{\measure}(\extractmc(\vartistc(\dsectionc))(\dbasepoint))
\end{equation}
The change in measurements $\dfunctc_{\measure}$ is defined by the developer as the symmetry between data and graphic that the artist is expected to preserve.

@@

\subsubsection{Addition: +}
\label{sec:artist:addition}
As illustrated in \autoref{eq:artist:operator}, data bundles can be combined by taking the disjoint  union of base spaces $\dbase^{a} \sqcup_{\dbase^c} \dbase^b$, where $\dbase^c$ is an overlap. When the fibers of each bundle are isomorphic $\dfiber^{a} \simeq \dfiber^{b}$, which we denote as $\dtotal$, this is analogous to adding more records to a dataset. We propose an addition operator that states that an artist that takes in a dataset can be constructed using artists that take as inputs subsets of the dataset
